Analysing Constraint Grammar with a SAT solver

### 2015-10-19 paper/lrec/implementation.tex
more text

In this section, we describe the implementation of the toolset.
We encode CG rules as a SAT problem, as in \cite{listenmaa_claessen2015}, and from the morphological lexicon, we get constraints for individual words. This allows us to create word sequences that will or will not trigger a given rule or rule set.

\paragraph{SAT encoding of CG}

Table~\ref{table:vars} shows the translation of the analyses to variables, and rules to clauses. The column ``Default rules'' refers to the rule that each word must have at least one analysis.
See \cite{listenmaa_claessen2015} for further description.

@@

Given a large morphological lexicon, we can restrict each individual word to contain only realistic ambiguities. We expand the lexicon, map each wordform to its possible analyses, and then ignore the actual wordforms. For instance, a Spanish morphological lexicon contains the entries \texttt{casa:<verb><sg><p3>} and \texttt{casa:<noun><sg>}, hence we know that a confusion between a 3rd person singular verb and a singular noun is possible. 
%The lexicon is not likely to contain a wordform that can be analysed both as an adjective and punctuation, so that kind of ambiguous word is never created.

>>>>

In this section, we describe the implementation of the toolset.
\cite{listenmaa_claessen2015} presents the SAT encoding of the rules, when applied to an existing text.
Since we want to test the rules in a way that doesn't depend on data, we create a \emph{symbolic sentence}: a word sequence which we can modify according to the properties we want to test.
These words are combinations of tags
In order to create words with real ambiguities, we use a morphological lexicon
This sentence consists of symbolic words, which we c
In addition, we use a morphological lexicon to create symbolic words that contain real ambiguities.

\paragraph{SAT encoding of CG rules}

A single CG rule is a template which creates a concrete clause for each instance where it is applied.
%Each potential analysis is represented as a variable, and the rules are translated into clauses that operate on those variables.
Table~\ref{table:vars} shows the rule \texttt{REMOVE v IF (-1 det)} applied to the text \emph{the bear}.
The analyses of the word forms are represented as variables, and
the default rule ensures that each word has at least one analysis.
%This makes \texttt{the\_det} true by default, therefore the implication must hold, and \texttt{bear\_v\_pl} will be removed.
For more details about the encoding, see \cite{listenmaa_claessen2015}.

%As shown in table~\ref{table:vars}, \texttt{bear\_n\_sg} represents the possibility that \emph{bear} is a noun, and \texttt{bear\_v\_pl} 
%The rules are translated into clauses; one for each concrete application.
%Table~\ref{table:vars} shows the translation of the analyses to variables, and rules to clauses. See \cite{listenmaa_claessen2015} for further description.

@@

The symbolic sentence consists of symbolic words.
They are combinations of tags, and they may be underspecified---similarly, a given CG rule may target any noun, a plural noun or the noun \emph{bear}.
Given a large morphological lexicon, we can restrict each individual word to contain only realistic ambiguities. We map each word form in the lexicon to its possible analyses, and then ignore the actual word forms. For instance, a Spanish morphological lexicon contains the entries \texttt{casa:casar<verb><sg><p3>} and \texttt{casa:casa<noun><sg>}, hence we know that a confusion between a 3rd person singular verb and a singular noun is possible. 
%The lexicon is not likely to contain a wordform that can be analysed both as an adjective and punctuation, so that kind of ambiguous word is never created.

### 2015-10-19 paper/lrec/lrec2016_listenmaa_claessen.tex
more text

\textbf{Analyses}                                                                                                 & \textbf{Variables}                                                          & \textbf{Default rules}                                    & \textbf{Constraint rules}                                                          \\ \hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}``\textless{}the\textgreater''\\$\qquad\quad$"the" det\end{tabular}}     & \multicolumn{1}{l|}{\texttt{the\_det}}                                      & \multicolumn{1}{l|}{\texttt{the\_det}}                    & \multicolumn{1}{l|}{\multirow{2}{*}{\texttt{the\_det $\Rightarrow$ $\neg$bear\_v\_pl}}} \\ \cline{1-3}
\begin{tabular}[c]{@{}l@{}}"\textless{}bear\textgreater"\\$\qquad\quad$"bear" n sg\\$\qquad\quad$"bear" v pl\end{tabular} & \texttt{\begin{tabular}[c]{@{}l@{}}bear\_n\_sg \\ bear\_v\_pl\end{tabular}} & \multicolumn{1}{l|}{\texttt{bear\_n\_sg $\vee$ bear\_v\_pl}} & \multicolumn{1}{l|}{}                                                              \\ \hline
\end{tabular}
\caption{SAT-encoding of \texttt{REMOVE v IF (-1 det)} applied to \emph{the bear}}
\label{table:vars}
\end{table*}

>>>>

\textbf{Analyses}
& \textbf{Variables}
& \textbf{Default rule: word has $\geq$1 analysis}                                    & \textbf{Constraint rules}                                                          \\ \hline
\multicolumn{1}{|l|}{\begin{tabular}[c]{@{}l@{}}``\textless{}the\textgreater''\\$\qquad\quad$"the" det\end{tabular}}     & \multicolumn{1}{l|}{\texttt{the\_det}}                                      & \multicolumn{1}{l|}{\texttt{the\_det}}                    & \multicolumn{1}{l|}{\multirow{2}{*}{\texttt{the\_det $\Rightarrow$ $\neg$bear\_v\_pl}}} \\ \cline{1-3}
\begin{tabular}[c]{@{}l@{}}"\textless{}bear\textgreater"\\$\qquad\quad$"bear" n sg\\$\qquad\quad$"bear" v pl\end{tabular} & \texttt{\begin{tabular}[c]{@{}l@{}}bear\_n\_sg \\ bear\_v\_pl\end{tabular}} & \multicolumn{1}{l|}{\texttt{bear\_n\_sg $\vee$ bear\_v\_pl}} & \multicolumn{1}{l|}{}                                                              \\ \hline
\end{tabular}
\caption{\texttt{REMOVE v IF (-1 det)} applied to \emph{the bear}}
\label{table:vars}
\end{table*}

### 2015-10-19 paper/lrec/problem_desc.tex
more text

Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin \cite{voutilainen2004}.
This helps the grammar writer to arrange the rules in appropriate sections, from safest and most effective to less certain and unusual. 
However, this method will not notice a missed opportunity or a grammar-internal conflict, nor suggest ways to improve.

@@

\cite{bick2013tuning} modifies the grammar automatically, by trying out different orderings of the rules, and changing their contexts to stricter or less strict.
Bick reports error reductions of 7--15\% compared to the original grammars.
However, the grammar writer will likely not know why exactly is the tuned grammar performing better.
% At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
% This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.

>>>>

Hand-annotated corpora are commonly used in the development of CGs, because they give immediate feedback whether a new rule increases or decreases accuracy, and by how large margin \cite{voutilainen2004}.
This helps the grammar writer to arrange the rules in appropriate sections, with safest and most effective rules coming first.
However, this method will not notice a missed opportunity or a grammar-internal conflict, nor suggest ways to improve.

@@

\cite{bick2013tuning} modifies the grammar automatically, by trying out different orderings of the rules, and changing their contexts to stricter or less strict.
Bick reports error reduction of 7--15\% compared to the original grammars.
As a downside, the grammar writer will likely not know why exactly is the tuned grammar performing better.
% At a certain point, the grammar gets so big that it is hard to keep track of all the rules and their interactions. \cite{bick2013tuning} tries out combinations of moving rules in different sections or removing them in total, and in parallel, making their contexts stricter or less strict. 
% This is a valuable tool, especially for grammars that are so big that it's hard to keep track of. Program can try all combinations whereas trying to make sense out of a huge set of rules would be hard for humans.
